A new machine-learning-based prediction of survival in patients with end-stage liver disease

### 2022-12-01 analysis/article.Rmd
fix: typos

  **Conclusion**:
      We provide a new machine-learning-based model of end-stage liver disease
      that incorporates synthesis and inflammatory markers and
      could improve the classical MELD score for 90-day survival prediction.

@@

Liver cirrhosis is the terminal result of the fibrotic remodeling
of liver tissue due to chronic damage.
This end-stage of liver disease is usually irreversible, and
the only available therapy is liver transplantation.
However, the shortage of grafts for transplantation from deceased donors
requires risk stratification and precise and fair allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine, and
the international normalized ratio (INR).
The MELD score has been extended by the sodium level
(MELD-Na score) because this was found to be an important additional
risk factor in liver cirrhosis [@kim2008; @optn_policies2021].
Recently MELD 3.0 was introduced adding a factor for female sex, albumin and
interactions between bilirubin and sodium and between albumin and creatinine.
The MELD was initially developed to predict the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
It was subsequently revalidated to predict mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score,
creatinine and INR are highly dependent on the laboratory methods used
[@trotter2004; @cholongitas2007].
In addition women are disadvantaged in MELD-Na [@sealock2022].
Patients with identical disease states can have very different MELD scores and
thus receive different priority levels on the liver transplantation waiting
list.
Furthermore, for acute-on-chronic liver failure, for example, the MELD score
often underestimates the mortality risk [@hernaez2020].

@@

This study aims to find clinical and laboratory values that improve on the risk
stratification for liver transplantation of the classical
MELD, MELD-Na, MELD 3.0 and MELD-Plus scores and that might be implemented as
part of the AMPEL clinical decision support system.

@@

MELD 3.0 was calculated as described in kim2022:

$$
\begin{aligned}
MELD3 = 1.33 (if female) &+
        4.56 &* \ln(bilirubin [mg/dL]) +  \\
        0.82 &* (137 - Na [mmol/L]) - 0.24 * (137 - Na [mmol/L]) * \ln(bilirubin [mg/dL]) + \\
        9.09 &* \ln(INR) + \\
        11.14 &* \ln(creatinine [mg/dL]) + 1.85 * (3.5 - albumin [g/dL]) - 1.83 * (3.5 - albumin [g/dL]) * \ln(creatinine [mg/dL]) + 6
\end{aligned}
$$
In contrast to MELD-Na the maximum allowed creatinine was 3.0 mg/dL.

@@

```{r tbl-observed-vs-expected-mortality, results = "asis", echo = FALSE, message = FALSE}
tbl_observed_vs_expected_mortality$ObservedMortality <-
    tbl_observed_vs_expected_mortality$ObservedMortality * 100
tbl_observed_vs_expected_mortality$ExpectedMortality <-
    tbl_observed_vs_expected_mortality$ExpectedMortality * 100
knitr::kable(
    cbind(
        row.names(tbl_observed_vs_expected_mortality),
        tbl_observed_vs_expected_mortality
    ),
    row.names = FALSE,
    col.names =
        c(
            "MELD-Na category",
            "Observed deaths (n)",
            "Expected deaths (n)",
            "Standardized mortality ratio (SMR)",
            "Observed mortality (%)",
            "Expected mortality (%)"
        ),
    caption = paste0(
        "Observed vs. MELD-Na-expected 90-day mortality. ",
        "MELD-Na mortality values are taken from @vanderwerken2021 and ",
        "the mean value was calculated for each MELD-Na category.",
        "All patients censored before day 90 are ignored for ",
        "the calculation of the MELD-Na-expected deaths. ",
        "SMR, Standardized mortality ratio = observed deaths/expected deaths."
    ),
    digits = 1
)
```

@@

In this retrospective analysis of `r n_analysed` consecutive patients
who were recruited during the evaluation process for liver transplantation at
University Hospital Leipzig, we developed a new model to predict
90-day survival probability. Our new model, AMELD,
incorporates synthesis and inflammatory markers and
could improve the prediction over previous published
MELD scores.

@@

The latter is an example where results from machine-learning can aid in
finding clinically meaningful patterns and guide further research.
Systemic inflammatory response syndrome is an
independent risk factor for poor outcomes in patients with cirrhosis
[@thabut2007].
However, neither MELD, MELD-Na nor MELD 3.0 took inflammation into account.
By contrast, IL-6 is the most important variable in our model. Even if IL-6 is
excluded, the algorithms select CRP as one of the most important variables,
which highlights the important role of inflammation in the prognosis of
patients with end-stage liver disease.
We and other researchers have already shown that adding inflammatory markers,
namely IL-6 or CRP, to the classical MELD score increases the accuracy of the
90-day mortality prediction [@remmler2018; @cervoni2016].

@@

In conclusion, we have developed a new model to predict the 90-day survival of
patients during the evaluation process for liver transplantation.
Our model, AMELD, extends the classical MELD predictors INR,
(direct) bilirubin, and cystatin C (instead of creatinine) by using the
synthesis parameters total protein and cholinesterase and,
most importantly, the inflammatory marker IL-6.
Using these six laboratory measurements, AMELD
could improve the prediction over previous published
MELD scores.
For wider adoption, a prospective multi-center study for model
recalibration and validation is needed.

>>>>

  **Conclusion**:
      We provide a new machine-learning-based model of end-stage liver disease
      that incorporates synthesis and inflammatory markers and
      may improve the classical MELD score for 90-day survival prediction.

@@

Liver cirrhosis is the terminal result of the fibrotic remodeling
of liver tissue due to chronic damage.
This end-stage of liver disease is usually irreversible, and
the only available therapy is liver transplantation.
However, the shortage of grafts for transplantation from deceased donors
requires risk stratification and precise and fair allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine, and
the international normalized ratio (INR).
The MELD score has been extended by the sodium level
(MELD-Na score) because this was found to be an important additional
risk factor in liver cirrhosis [@kim2008; @optn_policies2021].
Recently MELD 3.0 was introduced adding a factor for female sex, albumin and
interactions between bilirubin and sodium and between albumin and creatinine
[@kim2021].
The MELD was initially developed to predict the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
It was subsequently revalidated to predict mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score,
creatinine and INR are highly dependent on the laboratory methods used
[@trotter2004; @cholongitas2007].
In addition women are disadvantaged in MELD-Na [@sealock2022].
Patients with identical disease states can have very different MELD scores and
thus receive different priority levels on the liver transplantation waiting
list.
Furthermore, for acute-on-chronic liver failure, for example, the MELD score
often underestimates the mortality risk [@hernaez2020].

@@

This study aims to find clinical and laboratory values that improve on the risk
stratification for liver transplantation of the classical
MELD, MELD-Na, MELD 3.0, and MELD-Plus scores and that might be implemented as
part of the AMPEL clinical decision support system.

@@

MELD 3.0 was calculated as described in kim2021:

$$
\begin{aligned}
MELD3 = 1.33 (if female) &+
        4.56 &* \ln(bilirubin [mg/dL]) +  \\
        0.82 &* (137 - Na [mmol/L]) - 0.24 * (137 - Na [mmol/L]) * \ln(bilirubin [mg/dL]) + \\
        9.09 &* \ln(INR) + \\
        11.14 &* \ln(creatinine [mg/dL]) + 1.85 * (3.5 - albumin [g/dL]) - 1.83 * (3.5 - albumin [g/dL]) * \ln(creatinine [mg/dL]) + 6
\end{aligned}
$$
In contrast to MELD-Na the maximum allowed creatinine was set to 3.0 mg/dL.

@@

```{r tbl-observed-vs-expected-mortality, results = "asis", echo = FALSE, message = FALSE}
tbl_observed_vs_expected_mortality$ObservedMortality <-
    tbl_observed_vs_expected_mortality$ObservedMortality * 100
tbl_observed_vs_expected_mortality$ExpectedMortality <-
    tbl_observed_vs_expected_mortality$ExpectedMortality * 100
knitr::kable(
    cbind(
        row.names(tbl_observed_vs_expected_mortality),
        tbl_observed_vs_expected_mortality
    ),
    row.names = FALSE,
    col.names =
        c(
            "MELD-Na category",
            "Observed deaths (n)",
            "Expected deaths (n)",
            "Standardized mortality ratio (SMR)",
            "Observed mortality (%)",
            "Expected mortality (%)"
        ),
    caption = paste0(
        "Observed vs. MELD-Na-expected 90-day mortality. ",
        "MELD-Na mortality values are taken from @vanderwerken2021 and ",
        "the mean value was calculated for each MELD-Na category. ",
        "All patients censored before day 90 are ignored for ",
        "the calculation of the MELD-Na-expected deaths. ",
        "SMR, Standardized mortality ratio = observed deaths/expected deaths."
    ),
    digits = 1
)
```

@@

In this retrospective analysis of `r n_analysed` consecutive patients
who were recruited during the evaluation process for liver transplantation at
University Hospital Leipzig, we developed a new model to predict
90-day survival probability. Our new model, AMELD,
incorporates synthesis and inflammatory markers and
may improve the prediction over previous published
MELD scores.

@@

The latter is an example where results from machine-learning can aid in
finding clinically meaningful patterns and guide further research.
Systemic inflammatory response syndrome is an
independent risk factor for poor outcomes in patients with cirrhosis
[@thabut2007].
However, neither MELD, MELD-Na, nor MELD 3.0 took inflammation into account.
By contrast, IL-6 is the most important variable in our model. Even if IL-6 is
excluded, the algorithms select CRP as one of the most important variables,
which highlights the important role of inflammation in the prognosis of
patients with end-stage liver disease.
We and other researchers have already shown that adding inflammatory markers,
namely IL-6 or CRP, to the classical MELD score increases the accuracy of the
90-day mortality prediction [@remmler2018; @cervoni2016].

@@

In conclusion, we have developed a new model to predict the 90-day survival of
patients during the evaluation process for liver transplantation.
Our model, AMELD, extends the classical MELD predictors INR,
(direct) bilirubin, and cystatin C (instead of creatinine) by using the
synthesis parameters total protein and cholinesterase and,
most importantly, the inflammatory marker IL-6.
Using these six laboratory measurements, AMELD
may improve the prediction over previous published
MELD scores.
For wider adoption, a prospective multi-center study for model
recalibration and validation is needed.

### 2022-12-01 analysis/supplement.Rmd
fix: typos

```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0 and MELD-Plus7, as described in [@blanche2013]."}
tar_load(timeROC_MELD)
tar_load(timeROC_MELDNa)
tar_load(timeROC_MELD3)
tar_load(timeROC_MELDPlus7)
tar_load(timeROC_RCV)

@@

```{r roc-plot-wo-cb, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0 and MELD-Plus7, as described in [@blanche2013]. Identical to the figure above but without confidence bands."}
m <- lapply(m, function(x) { x$confint <- NULL;x })
plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(6)[-2])
```

>>>>

```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, and MELD-Plus7, as described in [@blanche2013]."}
tar_load(timeROC_MELD)
tar_load(timeROC_MELDNa)
tar_load(timeROC_MELD3)
tar_load(timeROC_MELDPlus7)
tar_load(timeROC_RCV)

@@

```{r roc-plot-wo-cb, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, and MELD-Plus7, as described in [@blanche2013]. Identical to the figure above but without confidence bands."}
m <- lapply(m, function(x) { x$confint <- NULL;x })
plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(6)[-2])
```
